cardinalR: Generating interesting high-dimensional data structures

### 2025-09-17 paper-cardinalR.Rmd
Updated gen_curvybranches

The `gen_curvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) c"} else { "Figure \\@ref(fig:branch-proj) c"})`). The underlying geometry lies in the first two dimensions, while the remaining $(p-2)$ dimensions contain Gaussian noise.
Branch $1$ is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1X_1 + X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane. Branch $2$ is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane. Branches $3$ to $k$ are added iteratively. Each new branch $i$ begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ to ensure connectivity with earlier branches, spans a unit-length interval on $X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure: $X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, \; s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}$. Here, $s_i$ is a sampled scale factor (slope-like term) controlling curvature, and $(x_{\text{start}}, y_{\text{start}})$ is the starting point from the existing structure. This construction yields $k$ spatially connected, nonlinear branches with varying curvature. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

>>>>

The `gen_curvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) c"} else { "Figure \\@ref(fig:branch-proj) c"})`). The underlying geometry lies in the first two dimensions, while the remaining $(p-2)$ dimensions contain Gaussian noise.

Branch $1$ is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1X_1 + X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane. Branch $2$ is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane. Branches $3$ to $k$ are then added iteratively. Each new branch $i$ begins at a randomly chosen starting point $(x_{\text{start}}, y_{\text{start}})$ from the existing structure, where the horizontal coordinate $x_{\text{start}}$ is restricted to $[-0.15, 0.15]$ to ensure connectivity with the initial branches. From this starting point, the branch extends over a unit-length interval $X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, with structure $X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, ; s_i \in {-2, -1.5, -1, -0.5, 0, 0.5, 1.5}$. Here, each branch $i$ is assigned a single scale factor $s_i$, sampled independently from the set, which controls the curvature of that branch. The parameter $(x_{\text{start}}, y_{\text{start}})$ is the starting point chosen from the existing structure. This construction yields $k$ spatially connected, nonlinear branches with varying curvature. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

### 2025-09-17 paper-cardinalR.tex
Updated gen_curvybranches

The \texttt{gen\_curvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in \(p\text{-}D\) (Figure \ref{fig:branch-proj} c). The underlying geometry lies in the first two dimensions, while the remaining \((p-2)\) dimensions contain Gaussian noise.
Branch \(1\) is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1X_1 + X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane. Branch \(2\) is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane. Branches \(3\) to \(k\) are added iteratively. Each new branch \(i\) begins at a randomly chosen point within a restricted horizontal band: \(X_{1} \in [-0.15, 0.15]\) to ensure connectivity with earlier branches, spans a unit-length interval on \(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), and has the structure: \(X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, \; s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}\). Here, \(s_i\) is a sampled scale factor (slope-like term) controlling curvature, and \((x_{\text{start}}, y_{\text{start}})\) is the starting point from the existing structure. This construction yields \(k\) spatially connected, nonlinear branches with varying curvature. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

A pyramid structure (Figure \ref{fig:pyr-proj}) represents data arranged around a central apex and base, useful for exploring how algorithms handle pointed or layered geometries in high-dimensional space. The functions provided allow users to generate pyramids with various base shapes---including rectangular, triangular, and star-shaped bases---with options for sharp or blunted apexes. Additionally, we offer functionality for creating pyramids containing structured holes, enabling the study of non-convex and sparse regions. These designs are valuable for testing the ability of dimension reduction and clustering algorithms to preserve and detect complex geometric features. Table \ref{tab:pyramid-tb-pdf} summarizes these functions.

@@

\paragraph{\texorpdfstring{\texttt{gen\_pyrrect()}}{gen\_pyrrect()}}\label{gen_pyrrect}

The \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a rectangular base (Figure \ref{fig:pyr-proj} a). The pyramid is embedded in \(p\text{-}D\), with a tip at height zero and base at height \(h\). The shape tapers linearly from the base dimensions \((l_x, l_y)\) to a smaller rectangular section with side lengths \((2r_t, 2r_t)\) at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

Let \(X_1, X_2, \dots, X_p\) denote the coordinates of the generated points. The final dimension \(X_p\) encodes the height of each point and is drawn from an exponential distribution capped at \(h\): \(X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)\). Let \(r_x(z)\) and \(r_y(z)\) denote the half-widths of the rectangular cross-section at height \(z\): \(r_x(z) = r_t + (l_x - r_t)z/h\), \(r_y(z) = r_t + (l_y - r_t)z/h\). The first three coordinates are then defined as: \(X_1 \sim U(-r_x(z),\ r_x(z))\), \(X_2 \sim U(-r_y(z),\ r_y(z))\), and \(X_3 \sim U(-r_x(z),\ r_x(z))\). For \(p > 3\), the remaining \(p - 4\) dimensions (i.e., \(X_4\) to \(X_{p-1}\)) are tapered toward zero with decreasing height and generated as: \(X_j \sim U(-0.1, 0.1)\left(1 - z/h\right), \quad \text{for } j = 4, \dots, p-1\). This tapering ensures that the structure narrows in all directions as it approaches the tip.

@@

\paragraph{\texorpdfstring{\texttt{gen\_pyrtri()}}{gen\_pyrtri()}}\label{gen_pyrtri}

The \texttt{gen\_pyrtri(n,\ p,\ h,\ l,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a triangular cross-section (Figure \ref{fig:pyr-proj} b). The structure is embedded in a \(p\text{-}D\), with the tip located at height \(0\) and the base at height \(h\). The triangle expands linearly in size from tip to base, with more points concentrated near the tip.

Let \(X_1, X_2, \dots, X_p\) denote the coordinates of the generated points. The final coordinate \(X_p\) encodes the height \(z\) of each point and is drawn from an exponential distribution capped at \(h\): \(X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)\). Let \(r(z)\) denote the scaling factor (distance from the origin to triangle vertices) at height \(z\): \(r(z) = r_t + (l - r_t)z/h\).

A point in the triangle at height \(z\) is generated using barycentric coordinates \((u, v)\) to ensure uniform sampling within the triangular cross-section: \(u, v \sim U(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v\). The first three coordinates (triangle plane) are then: \(X_1 = r(z)(1 - u - v)\), \(X_2 = r(z)u\), and \(X_3 = r(z)v\). For dimensions \(j = 4, \dots, p-1\), values are tapered linearly toward zero near the tip: \(X_j \sim U(-0.1, 0.1)\left(1 - z/h\right)\).

@@

\paragraph{\texorpdfstring{\texttt{gen\_pyrstar()}}{gen\_pyrstar()}}\label{gen_pyrstar}

The \texttt{gen\_pyrstar(n,\ p,\ h,\ rb)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-shaped structure with a hexagonal base (Figure \ref{fig:pyr-proj} c). The pyramid extends vertically from height \(z = 0\) (tip) to height \(z = h\) (base), embedded in a \(p\)-dimensional space. The distribution concentrates more points near the base, and the hexagonal spread tapers toward the tip.

Let \(X_1, X_2, \dots, X_p\) denote the coordinates of the generated points. The final coordinate \(X_p\) encodes the height \(z\): \(X_p = z \sim U(0, h)\).

The radius at height \(z\) scales linearly from zero (tip) to the base radius \(r_b\): \(r(z) = r_b\left(1 - z/h\right)\).

Each point is placed within a regular hexagon in the plane \((X_1, X_2)\), using a randomly chosen hexagon sector angle \(\theta \in \{0, \pi/3, 2\pi/3, \pi, 4\pi/3, 5\pi/3\}\) and a uniformly random radial scaling factor: \(\theta \sim \text{Uniform sample from 6 hexagon angles}\),
\(r_{\text{point}} \sim \sqrt{U(0, 1)}\). Then, the first two coordinates are: \(X_1 = r(z)r_{\text{point}}\cos(\theta)\), and \(X_2 = r(z)r_{\text{point}}\sin(\theta)\). For dimensions \(j = 3, \dots, p-1\), values are tapered toward zero near the tip: \(X_j \sim U(-0.1, 0.1)\left(1 - z/h \right)\).

>>>>

The \texttt{gen\_curvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in \(p\text{-}D\) (Figure \ref{fig:branch-proj} c). The underlying geometry lies in the first two dimensions, while the remaining \((p-2)\) dimensions contain Gaussian noise.

Branch \(1\) is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1X_1 + X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane. Branch \(2\) is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane. Branches \(3\) to \(k\) are then added iteratively. Each new branch \(i\) begins at a randomly chosen starting point \((x_{\text{start}}, y_{\text{start}})\) from the existing structure, where the horizontal coordinate \(x_{\text{start}}\) is restricted to \([-0.15, 0.15]\) to ensure connectivity with the initial branches. From this starting point, the branch extends over a unit-length interval \(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), with structure \(X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, ; s_i \in {-2, -1.5, -1, -0.5, 0, 0.5, 1.5}\). Here, each branch \(i\) is assigned a single scale factor \(s_i\), sampled independently from the set, which controls the curvature of that branch. The parameter \((x_{\text{start}}, y_{\text{start}})\) is the starting point chosen from the existing structure. This construction yields \(k\) spatially connected, nonlinear branches with varying curvature. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

A pyramid structure (Figure \ref{fig:pyr-proj}) represents data arranged around a central apex and base, useful for exploring how algorithms handle pointed or layered geometries in high-dimensional space. The functions provided allow users to generate pyramids with rectangular, triangular, and star-shaped bases, and sharp or blunted apexes. Additionally, similar to the cube generator, it is possible to create a pyramids containing structured holes, enabling the study of non-convex and sparse regions. Table \ref{tab:pyramid-tb-pdf} summarizes these functions.

Let \(X_1, \dots, X_p\) denote the coordinates of the generated points. For the rectangular and triangular based pyramid generator functions, the final dimension, \(X_p\), encodes the height of each point and is drawn from an exponential distribution capped at the maximum height \(h\). That is, \[X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)\]. This distribution creates a natural skew toward smaller height values, resulting in a denser concentration of points near the pyramid's apex. For the star-shaped base pyramid, the final dimension is drawn from a uniform distribution. That is, \(X_p = z \sim U(0, h)\).

The remaining dimensions are based on the specific pyramid shape. For the rectangular based pyramid, \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} (Figure \ref{fig:pyr-proj} a), let \(r_x(z)\) and \(r_y(z)\) denote the half-widths of the rectangular cross-section at height \(z\). That is, \(r_x(z) = r_t + (l_x - r_t)z/h\), \(r_y(z) = r_t + (l_y - r_t)z/h\). The first three coordinates are then defined as: \(X_1 \sim U(-r_x(z),\ r_x(z))\), \(X_2 \sim U(-r_y(z),\ r_y(z))\), and \(X_3 \sim U(-r_x(z),\ r_x(z))\).

For the triangular based pyramid, \texttt{gen\_pyrtri(n,\ p,\ h,\ l,\ rt)} (Figure \ref{fig:pyr-proj} b), let \(r(z)\) denote the scaling factor (distance from the origin to triangle vertices) at height \(z\). That is, \(r(z) = r_t + (l-r_t)z/h\). A point in the triangle at height \(z\) is generated using barycentric coordinates \((u, v)\) to ensure uniform sampling within the triangular cross-section: \(u, v \sim U(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v\). The first three coordinates (triangle plane) are then: \(X_1 = r(z)(1 - u - v)\), \(X_2 = r(z)u\), and \(X_3 = r(z)v\).

For the star based pyramid, \texttt{gen\_pyrstar(n,\ p,\ h,\ rb)} (Figure \ref{fig:pyr-proj} c), let the radius at height \(z\), \(r(z)\), be such that the radius scales linearly from zero (tip) to the base radius \(r_b\). That is, \(r(z) = r_b\left(1 - z/h\right)\).

Each point is placed within a regular hexagon in the plane \((X_1, X_2)\), using a randomly chosen hexagon sector angle \(\theta \in \{0, \pi/3, 2\pi/3, \pi, 4\pi/3, 5\pi/3\}\) and a uniformly random radial scaling factor: \(\theta \sim \text{Uniform sample from 6 hexagon angles}\),
\(r_{\text{point}} \sim \sqrt{U(0, 1)}\). Then, the first two coordinates are: \(X_1 = r(z)r_{\text{point}}\cos(\theta)\), and \(X_2 = r(z)r_{\text{point}}\sin(\theta)\).

For all pyramid shapes, if \(p > 3\), the remaining \(p - 4\) dimensions (i.e., \(X_4\) to \(X_{p-1}\)) are tapered toward zero with decreasing height and generated as: \(X_j \sim U(-0.1, 0.1)\left(1 - z/h\right), \quad \text{for } j = 4, \dots, p-1\). This tapering ensures that the structure narrows in all directions as it approaches the tip.

@@

@@

@@
